Privacy Calculus Experiment

### 2023-04-12 manuscript.RmD
- Language updates

---
title             : "Likes or Dislikes, Gratifications or Concerns? Analyzing Popularity Cues and Privacy Calculus in Online Communication"

@@

abstract: |
    According to the privacy calculus model, online communication is mainly determined by the two factors of privacy concerns and expected benefits. This paper analyzes how the privacy calculus is affected by popularity cues, namely like and dislike buttons. In addition, the privacy calculus is extended by elaborating on the process of comparing benefits and drawbacks, termed privacy deliberation, and by investigating the role of trust and self-efficacy. To this end, a preregistered one-week field experiment with 590 participants was conducted. Participants were randomly distributed to three different websites, on which they discussed a current political topic. The websites featured either (a) like buttons, (b) both like and dislike buttons, or (c) no like or dislike buttons. The exploratory analyses revealed, contrary to what was expected, that implementing both like and dislike buttons decreased online communication. As predicted by the privacy calculus model, expected benefits and perceived privacy concerns substantially affected communcation. Actively deliberating about privacy by comparing benefits and drawbacks decreased communication, while experiencing self-efficacy and trust strongly increased communication.

@@

<!-- Understanding why people share personal information online is important.  -->
According to the privacy paradox, the way users share information online is erratic [@barnesPrivacyParadoxSocial2006]:
<!-- Personal beliefs, privacy concerns, or behavioral attitudes do not affect online behavior. -->
<!-- [for an overview of the privacy paradox, see @dienlinPrivacyParadoxAus2019].  -->
People communicate vast amounts of personal information online, although they are concerned about their privacy [@taddickenUsesPrivacyOnline2011].
However, despite its popularity in both academic research [@dienlinPrivacyParadoxAus2019] and the public press [@newyorkpublicradioPrivacyParadox2018], empirical support for the privacy paradox is weak [@baruhOnlinePrivacyConcerns2017; @dienlinDoesPrivacyParadox2021]. 
<!-- , which shows that privacy concerns are indeed related to communication online.  -->
<!-- Rather than further pursuing the privacy paradox,  -->
Research is therefore increasingly building on the privacy calculus model, which states that communication online can be explained successfully by means of perceived risks and expected benefits [@bolUnderstandingEffectsPersonalization2018b; @krasnovaOnlineSocialNetworks2010; @meierPrivacyCalculusRevisited2022].
<!-- Several studies have shown that experiencing more privacy risks or concerns is related to sharing less information online [@bolUnderstandingEffectsPersonalization2018b; @meierPrivacyCalculusRevisited2022], while expecting more benefits is related to sharing more information online [@heirmanPredictingAdolescentsDisclosure2013; @koohikamaliInvestigationDynamicModel2019]. -->

Although the privacy calculus has gained momentum in academic research, several important questions remain unanswered. 
First, we still know little about how the privacy calculus and online communication can be affected by external influences [@bazarovaIntegrationIndividualisticNetworked2020; @braunlichLinkingLooseEnds2020].
For example, does the existence of popularity cues such as like and dislike buttons affect the privacy calculus and online communication?
<!-- I therefore analyze whether the privacy calculus is affected by the affordances of a website.  -->
<!-- Specifically, I investigate whether *popularity cues* such as like and dislike buttons affect the privacy calculus and whether they foster communication online. -->
Second, the privacy calculus' empirical foundation needs to be improved [@knijnenburgDeathPrivacyCalculus2017]. 
To date, much research on the privacy calculus builds on (a) self-reports of behavior [@krasnovaOnlineSocialNetworks2010], (b) vignette approaches [@bolUnderstandingEffectsPersonalization2018b], or (c) one-shot experiments in the lab [@treptePrivacyCalculusContextualized2020]. 
Can the privacy calculus be replicated in an authentic and more long-term setting using actual behavioral data?
Third, research on the privacy calculus so far did not measure the actual process of weighing pros and cons [@knijnenburgDeathPrivacyCalculus2017], 
<!-- According to critics , showing that both concerns and gratifications correlate with communication behavior online is not sufficient evidence for an explicit weighing process.  -->
<!-- In this study I will thus focus on the explicit process of comparing pros and cons  -->
which I will do in this study by elaborating on the privacy deliberation process. 
Fourth, the privacy calculus model is criticized for over-emphasizing rationality [@knijnenburgDeathPrivacyCalculus2017].
Hence, I extend the privacy calculus in analyzing how trust and self-efficacy, two more psychological variables, affect communication.
<!-- Third, in this study I approach the privacy calculus from the theoretical perspective of _bounded rationality_.  -->
<!-- It is likely that other factors next to risks and benefits also determine behavior. -->
<!-- I therefore extend the privacy calculus model theoretically by investigating the role and interplay of trust and self-efficacy. -->

@@

The privacy calculus analyzes why people communicate online.
When are people willing to engage in a conversation?
<!-- Do they express their opinion?  -->
It builds on the calculus of behavior [@lauferPrivacyConceptSocial1977], which states that people are weighing risks and benefits before communicating.
<!-- In communicating online, people share much information about themselves. -->
Communication is closely related to privacy, which is defined as a "voluntary and temporary withdrawal of a person from the general society" [@westinPrivacyFreedom1967, p. 7].
People regulate their privacy by deciding what and what not to communicate [@dienlinPrivacyProcessModel2014].
Disclosing information can be risky, and if people are more concerned about their privacy 
<!-- they should be less likely to communicate information. -->
<!-- Almost all types of communication allow for meaningful inferences about a person, even when we _don't_ communicate [@watzlawickPragmaticsHumanCommunication2011].  -->
<!-- However, the recent years have illustrated vividly how much additional insights we can get about a person, simply by analyzing their digital traces of communication or their meta-data [@kosinskiPrivateTraitsAttributes2013].  -->
<!-- The more we communicate, the more we self-disclose. -->
<!-- Online privacy concerns represent "[...] how much an individual is motivated to focus on their control over a voluntary withdrawal from other people or societal institutions on the Internet, accompanied by an uneasy feeling that their privacy might be threatened" [@dienlinLongitudinalAnalysisPrivacy2021, p. 4].  -->
<!-- Notably, I don't assume that this weighing process is flawless or perfectly rational. -->
<!-- Instead, I understand the privacy calculus from the perspective of _bounded rationality_ [@simonBoundedRationality1990], which has three main tenets:  -->
<!-- "(1) humans are cognitively constrained;  -->
<!-- (2) these constraints impact decision making;  -->
<!-- and (3) difficult problems reveal the constraints and highlight their significance." [@bendorBoundedRationality2015, p. 1303] -->
<!-- Crucially, although bounded rationally upholds that human behavior is not perfectly logical, this does not meant that it is illogical [@gigerenzerBoundedRationalityAdaptive2002]. -->
<!-- Instead, it makes more sense to consider rationality a continuum. -->
<!-- Humans are still trying to optimize the outcomes of their behavior according to their own best interests or values. -->
<!-- It is only that their capacity to do so is bounded. -->
<!-- Transferred to the context of online privacy, it is by now well known that several irregularities and inconsistencies between concerns and communication behavior exist. -->
<!-- These differences stem from, for example, information asymmetries, present bias, intangibility, illusory control, or herding [@acquistiSecretsLikesDrive2020]. -->
<!-- Indeed, If people experience more concerns,  -->
they are less likely to communicate online [@baruhOnlinePrivacyConcerns2017; @dienlinExtendedPrivacyCalculus2016a; @krasnovaOnlineSocialNetworks2010; @masurUnderstandingEffectsConceptual2023; @meierPrivacyCalculusRevisited2022].
According to a meta-analysis by @baruhOnlinePrivacyConcerns2017, the effect is small but significant (*r* = -.13).
<!-- At the same time, _on average_ people do behave according to their interests, respond to incentives, or actively manage their privacy [@baruhOnlinePrivacyConcerns2017; @braunlichLinkingLooseEnds2020; @dienlinExtendedPrivacyCalculus2016a; @soloveMythPrivacyParadox2020]. -->

@@

Outweighing concerns, the most relevant drivers of online communication are expected gratifications [@dienlinExtendedPrivacyCalculus2016a; @bolUnderstandingEffectsPersonalization2018b; @kezerGettingPrivacyCalculus2022].
People are happy to trade in parts of their privacy, when they can get something more valuable in return [@lauferPrivacyConceptSocial1977].
In online communication, the most important benefits include social support [@krasnovaOnlineSocialNetworks2010], social capital [@ellisonNegotiatingPrivacyConcerns2011], entertainment [@dhirUnderstandingRelationshipIntensity2017], information-seeking [@whitingWhyPeopleUse2013], and self-presentation [@minHowArePeople2015].
<!-- As a result, we expect a moderate relationship. -->

H1: People who are more concerned about their privacy than others are less likely to communicate actively on a website.

@@

Although privacy calculus implies that people explicitly compare benefits and drawbacks before communicating online, research has neglected this aspect [@knijnenburgDeathPrivacyCalculus2017].
Only observing that privacy concerns or expected gratifications and communication online are related does not prove that an explicit weighing process took place [@knijnenburgDeathPrivacyCalculus2017]. 
Instead, we need to analyze if and how much people actively deliberate about their privacy, and whether doing so influences their decision to communicate.
<!-- We can understand the privacy calculus from two perspectives. -->
<!-- First, is the communication behavior aligned with people's privacy concerns and expected benefits?  -->
<!-- Second, is the communication process implicit or explicit? -->
<!-- Here, I suggest that the privacy calculus should be discussed in light of dual process theories, which state that people either deliberately, explicitly, and centrally take decisions, or instead do so automatically, implicitly, and peripherally [@kahnemanThinkingFastSlow2011; @pettyCommunicationPersuasionCentral1986].  -->
<!-- Accordingly, privacy calculus would assume that people, when it comes to disclosing, engage in a central processing.  -->
Self-disclosure theory [@omarzuDisclosureDecisionModel2000;  @altmanPrivacyConceptualAnalysis1976] suggests that if the benefits to be attained from communication are attractive, deciding whether or not to communicate is a "conscious and deliberative process"  [@omarzuDisclosureDecisionModel2000, p. 183].
I hence introduce and investigate a novel concept termed _privacy deliberation_.
Privacy deliberation captures the extent to which individuals explicitly compare risks and benefits before communicating with others.
How could deliberating on one's privacy affect communication?
On the one hand, deliberating about privacy could reduce subsequent communication.
Refraining from communication---the primary means of connecting with others [@altmanPrivacyConceptualAnalysis1976]---likely requires some active and deliberate restraint [@omarzuDisclosureDecisionModel2000]. 
This is especially true for social media, which are designed to foster communication and participation [@ellisonSocialNetworkSite2015; @masurBehavioralContagionSocial2021].
Actively thinking about whether communicating is worthwhile might be the first step not to participate.
On the other hand, deliberating about privacy might also increase communication.
In light of the benefits mentioned above, a person might conclude that communication is beneficial, thereby legitimating communication [@krasnovaOnlineSocialNetworks2010]. 
<!-- I therefore formulate the following research question: -->
<!-- Deliberation could represent some kind of inner consent, providing additional affirmation. -->
<!-- Alternatively, it could be that deliberation functions as a moderator. -->
<!-- For example, if people actively deliberate about whether or not to disclose, this might reinforce the effect of concerns or gratifications. -->
<!-- Reflecting about the pros and cons of communication might concerns and gratifications more salient.  -->
<!-- Alternatively, it could also be that deliberating decreases the effects, for example because apparent gratifications are considered more critically, and maybe loose their appeal. -->

RQ1: Do people who deliberate more actively whether they should communicate, communicate more or less online?

<!-- RQ2: Do people who deliberate more actively whether they should communicate, show larger or smaller relations between concerns, gratifications and communication behavior? -->
This weighing process is not flawless or perfect.
It is therefore useful to understand the privacy calculus from a perspective of _bounded rationality_ [@simonBoundedRationality1990].
Bounded rationality states that "(1) humans are cognitively constrained; (2) these constraints impact decision making; and (3) difficult problems reveal the constraints and highlight their significance." [@bendorBoundedRationality2015, p. 1303].
But although human behavior is partly irrational, bounded rationality does not state that it is completely irrational [@gigerenzerBoundedRationalityAdaptive2002].
Rationality is a continuum. 
<!-- Instead, it makes more sense to consider rationality a continuum. -->
<!-- Humans are still trying to optimize the outcomes of their behavior according to their own best interests or values. -->
<!-- It is only that their capacity to do so is bounded. -->
In the context of online privacy, for example, rationality is impeded by information asymmetries, presence bias, intangibility, illusory control, or herding [@acquistiSecretsLikesDrive2020].
It follows that to provide a more complete picture, additional factors next to gratifications and concerns should also explain communication.
Factors less focused on rationality, but more on psychosocial aspects.
<!-- Communication online often takes place in situations where information is limited or obscure. -->

@@

In situations where people lack experience or competence, a relevant variable to understand behavior is _trust_.
Especially in online contexts, users often cannot control the environment or the way their information is handled [@acquistiSecretsLikesDrive2020; @braunlichLinkingLooseEnds2020].
Trust plays a key role in online communication [@metzgerPrivacyTrustDisclosure2004]. 
Trust either captures "_specific_ beliefs dealing primarily with the integrity, benevolence, and ability of another party" [@gefenTrustTAMOnline2003, p. 55, emphasis added] or a "_general_ belief that another party can be trusted" [@gefenTrustTAMOnline2003, p.55, emphasis added]. 
<!-- Whereas specific trust focuses on the causes of trust, general trust emphasizes the experience of trust. -->
In online contexts, there are different _targets_ of trust, including (a) the information system, (b) the provider, (c) the Internet, and (d) the community of other users [@sollnerWhyDifferentTrust2016]. 
People who put more trust in the providers of networks, for example, disclose more personal information [@liEmpiricalStudiesOnline2011]. 
<!-- @gefenTrustTAMOnline2003 prioritize specific trust (p. 60).  -->
<!-- Because the targets can be largely different, it is often recommended to analyze them individually. -->

@@

How are the privacy calculus and communication affected by the context, the digital infrastructure? 
First, it makes sense not to focus on specific features of particular websites, which change and become obsolete [@foxDistinguishingTechnologiesSocial2017]. 
Instead, it is recommended focusing on the underlying latent structures by analyzing so-called _affordances_ [@ellisonSocialNetworkSite2015; @foxDistinguishingTechnologiesSocial2017].  
Affordances are mental representations of how objects might be used.
They emphasize that it is not the _objective features_ that determine behavior, but rather our _subjective perceptions_ [@gibsonEcologicalApproachVisual2015].
<!-- ; as such, they are by definition subjective.  -->
There is an ongoing debate on what exactly defines an affordance [@evansExplicatingAffordancesConceptual2017].
Whereas @evansExplicatingAffordancesConceptual2017 propose three affordances for mediated communication (i.e., anonymity, persistence, and visibility), @foxDistinguishingTechnologiesSocial2017 suggest 10 affordances for SNSs alone (i.e., accessibility, bandwidth, social presence, privacy, network association, personalization, persistence, editability, conversation control, and anonymity).
<!-- The privacy calculus states that both benefits and costs determine behavior. -->
Popularity cues such as like and dislike buttons, which are of interest here, are categorized as "paralinguistic digital affordances" [@carrPredictingThresholdPerceived2018, p. 142].

@@

Paralinguistic digital affordances and specifically popularity cues have been shown to affect behavior [@kramerMasteringChallengeBalancing2020; @masurBehavioralContagionSocial2021; @treptePrivacyCalculusContextualized2020].
Online comments that already have several dislikes are more likely to receive further dislikes [@muchnikSocialInfluenceBias2013]. 
When users disagree with a post, they are more likely to click on a button labeled _respect_ compared to a button labeled _like_ [@stroudRecommendRespectAltering2017].
<!-- In light of their potentially strong negative effects, most major social networking sites did not implement dislike buttons.  -->
<!-- At the time of writing, only a handful of websites have (partially) implemented it (e.g., youtube, stackexchange, or reddit). -->
Popularity cues likely impact the privacy calculus via two underlying theoretical mechanisms [@kramerMasteringChallengeBalancing2020]:
First, the _mere presence_ of popularity cues might affect whether people are willing to disclose. 
Being able to attract likes might motivate users to communicate, while the mere option to receive dislikes might intimidate others.
Second, _actually receiving_ likes or dislikes might additionally affect subsequent behavior, thereby reinforcing the process.

Likes are positive and represent the positivity bias of social media [@schreursExposurePositivityBias2022]. 
Receiving a like online is similar to receiving a compliment offline [@carrPredictingThresholdPerceived2018; @sumnerFunctionalApproachFacebook2017].
Like-buttons afford and emphasize a _gain frame_ [@rosoffHeuristicsBiasesCyber2013].
These gains can be garnered only through active participation. 
Because like buttons emphasize positive outcomes, it is likely that concerns decrease. 
In situations where there is more to win, people might also more actively deliberate about whether or not to communicate. 
Dislike represent a punishment, introducing a _loss frame_. 
<!-- Although most communication emphasizes positive aspects, the Internet is also replete with spite, envy, and arguments. -->
Websites featuring both like _and_ dislike buttons should be more ambivalent compared to websites without popularity cues.
<!-- Especially in online contexts, short-term gains often outweigh long-term losses [@acquistiSecretsLikesDrive2020]. -->
<!-- Having both like and dislike buttons might still lead to more gratifications and communication. -->
Privacy concerns should not be reduced anymore:
People who are more concerned about their privacy are also more shy and risk averse [@dienlinPsychologyPrivacyAnalyzing2017].
Implementing the dislike button might therefore increase privacy concerns, thereby canceling out the positive effects of the like button. 
<!-- And because there is more at stake, participants should deliberate even more whether or not to disclose.  -->

@@

This manuscript features a [companion website](https://XMtRa.github.io/like_dislike), which includes the data, research materials, analyses scripts, and a reproducible version of this manuscript (see https://XMtRa.github.io/like_dislike).
The hypotheses, sample size, research materials, analyses, and exclusion criteria were preregistered (see https://osf.io/a6tzc/?view_only=5d0ef9fe5e1745878cd1b19273cdf859). 
In some cases, the pre-registered approach needed to be changed (see [companion website](https://XMtRa.github.io/like_dislike)). 
Analyses that were not preregistered are reported as exploratory analyses.

>>>>

---
title             : "Likes or Dislikes, Gratifications or Concerns? Analyzing the Effects of Popularity Cues on the Privacy Calculus in Online Communication"

@@

abstract: |
    According to the privacy calculus model, online communication is determined by two factors, privacy concerns and expected benefits. How is the privacy calculus affected by external affordances, for example popularity cues such as like and dislike buttons? And should the privacy calculus model be extended by integrating more socio-psychological measures such as trust, self-efficacy, and privacy deliberation? To this end, a preregistered one-week field experiment with 590 participants was conducted. Participants were randomly distributed to three different websites, on which they discussed a current political topic. The websites included either (a) like buttons, (b) both like and dislike buttons, or (c) no like or dislike buttons. Because of multicollinearity and non-normal distribution of the outcome, the preregistered analyses needed to be updated. The results revealed that implementing both like and dislike buttons decreased online communication by 44%. As predicted by the privacy calculus model, expected benefits and perceived privacy concerns substantially affected communcation. Actively deliberating about privacy by comparing benefits and drawbacks decreased communication, while experiencing self-efficacy and trust strongly increased communication.

@@

<!-- Understanding why people share personal information online is important.  -->
According to the privacy paradox, the way users share information online is erratic [@barnesPrivacyParadoxSocial2006]:
<!-- Personal beliefs, privacy concerns, or behavioral attitudes do not affect online behavior. -->
<!-- [for an overview of the privacy paradox, see @dienlinPrivacyParadoxAus2019].  -->
People communicate vast amounts of personal information online, although they are concerned about their privacy [@taddickenUsesPrivacyOnline2011].
However, despite its popularity in both academic research [@dienlinPrivacyParadoxAus2019] and the public press [@newyorkpublicradioPrivacyParadox2018], empirical support for the privacy paradox is weak [@baruhOnlinePrivacyConcerns2017; @dienlinDoesPrivacyParadox2021]. 
<!-- , which shows that privacy concerns are indeed related to communication online.  -->
<!-- Rather than further pursuing the privacy paradox,  -->
Research is therefore increasingly building on the privacy calculus model, which states that communication online can be explained successfully by perceived risks and expected benefits [@bolUnderstandingEffectsPersonalization2018b; @krasnovaOnlineSocialNetworks2010; @meierPrivacyCalculusRevisited2022].
<!-- Several studies have shown that experiencing more privacy risks or concerns is related to sharing less information online [@bolUnderstandingEffectsPersonalization2018b; @meierPrivacyCalculusRevisited2022], while expecting more benefits is related to sharing more information online [@heirmanPredictingAdolescentsDisclosure2013; @koohikamaliInvestigationDynamicModel2019]. -->

Although the privacy calculus has gained momentum in academic research, several important questions remain unanswered. 
First, we still know little about how the privacy calculus and online communication can be affected by external affordances [@bazarovaIntegrationIndividualisticNetworked2020; @braunlichLinkingLooseEnds2020].
For example, do popularity cues such as like and dislike buttons affect the privacy calculus and online communication?
<!-- I therefore analyze whether the privacy calculus is affected by the affordances of a website.  -->
<!-- Specifically, I investigate whether *popularity cues* such as like and dislike buttons affect the privacy calculus and whether they foster communication online. -->
Second, the privacy calculus' empirical foundation needs to be improved [@knijnenburgDeathPrivacyCalculus2017]. 
To date, much research on the privacy calculus builds on (a) self-reports of behavior [@krasnovaOnlineSocialNetworks2010], (b) vignette approaches [@bolUnderstandingEffectsPersonalization2018b], or (c) one-shot experiments in the lab [@treptePrivacyCalculusContextualized2020]. 
Can the privacy calculus be replicated in more authentic and long-term settings using actual behavioral data?
Third, research on the privacy calculus so far did not measure the actual process of weighing pros and cons [@knijnenburgDeathPrivacyCalculus2017], 
<!-- According to critics , showing that both concerns and gratifications correlate with communication behavior online is not sufficient evidence for an explicit weighing process.  -->
<!-- In this study I will thus focus on the explicit process of comparing pros and cons  -->
which I do in this study by elaborating on what I will call the the privacy deliberation process. 
Fourth, the privacy calculus model is criticized for over-emphasizing rationality [@knijnenburgDeathPrivacyCalculus2017].
Hence, I extend the privacy calculus in analyzing how trust and self-efficacy, two more socio-psychological variables, affect communication.
<!-- Third, in this study I approach the privacy calculus from the theoretical perspective of _bounded rationality_.  -->
<!-- It is likely that other factors next to risks and benefits also determine behavior. -->
<!-- I therefore extend the privacy calculus model theoretically by investigating the role and interplay of trust and self-efficacy. -->

@@

The privacy calculus analyzes why people communicate online.
When are we willing to engage in a conversation?
<!-- Do they express their opinion?  -->
It builds on the calculus of behavior [@lauferPrivacyConceptSocial1977] and states that people are weighing risks and benefits before communicating.
<!-- In communicating online, people share much information about themselves. -->
Communication is closely related to privacy, which is defined as a "voluntary and temporary withdrawal of a person from the general society" [@westinPrivacyFreedom1967, p. 7].
People regulate their privacy by deciding what and what not to communicate [@dienlinPrivacyProcessModel2014].
Disclosing information can be risky, and if people are more concerned about their privacy 
<!-- they should be less likely to communicate information. -->
<!-- Almost all types of communication allow for meaningful inferences about a person, even when we _don't_ communicate [@watzlawickPragmaticsHumanCommunication2011].  -->
<!-- However, the recent years have illustrated vividly how much additional insights we can get about a person, simply by analyzing their digital traces of communication or their meta-data [@kosinskiPrivateTraitsAttributes2013].  -->
<!-- The more we communicate, the more we self-disclose. -->
<!-- Online privacy concerns represent "[...] how much an individual is motivated to focus on their control over a voluntary withdrawal from other people or societal institutions on the Internet, accompanied by an uneasy feeling that their privacy might be threatened" [@dienlinLongitudinalAnalysisPrivacy2021, p. 4].  -->
<!-- Notably, I don't assume that this weighing process is flawless or perfectly rational. -->
<!-- Instead, I understand the privacy calculus from the perspective of _bounded rationality_ [@simonBoundedRationality1990], which has three main tenets:  -->
<!-- "(1) humans are cognitively constrained;  -->
<!-- (2) these constraints impact decision making;  -->
<!-- and (3) difficult problems reveal the constraints and highlight their significance." [@bendorBoundedRationality2015, p. 1303] -->
<!-- Crucially, although bounded rationally upholds that human behavior is not perfectly logical, this does not meant that it is illogical [@gigerenzerBoundedRationalityAdaptive2002]. -->
<!-- Instead, it makes more sense to consider rationality a continuum. -->
<!-- Humans are still trying to optimize the outcomes of their behavior according to their own best interests or values. -->
<!-- It is only that their capacity to do so is bounded. -->
<!-- Transferred to the context of online privacy, it is by now well known that several irregularities and inconsistencies between concerns and communication behavior exist. -->
<!-- These differences stem from, for example, information asymmetries, present bias, intangibility, illusory control, or herding [@acquistiSecretsLikesDrive2020]. -->
<!-- Indeed, If people experience more concerns,  -->
they are less likely to communicate online [@baruhOnlinePrivacyConcerns2017; @dienlinExtendedPrivacyCalculus2016a; @krasnovaOnlineSocialNetworks2010; @masurUnderstandingEffectsConceptual2023; @meierPrivacyCalculusRevisited2022].
According to a meta-analysis by @baruhOnlinePrivacyConcerns2017, the effect is small but significant (*r* = -.13).
<!-- At the same time, _on average_ people do behave according to their interests, respond to incentives, or actively manage their privacy [@baruhOnlinePrivacyConcerns2017; @braunlichLinkingLooseEnds2020; @dienlinExtendedPrivacyCalculus2016a; @soloveMythPrivacyParadox2020]. -->

@@

Outweighing concerns, the most relevant drivers of online communication are expected gratifications [@dienlinExtendedPrivacyCalculus2016a; @bolUnderstandingEffectsPersonalization2018b; @kezerGettingPrivacyCalculus2022].
People are happy to trade in parts of their privacy when they can receive something more valuable in return [@lauferPrivacyConceptSocial1977].
In online communication, the most important benefits include social support [@krasnovaOnlineSocialNetworks2010], social capital [@ellisonNegotiatingPrivacyConcerns2011], entertainment [@dhirUnderstandingRelationshipIntensity2017], information-seeking [@whitingWhyPeopleUse2013], and self-presentation [@minHowArePeople2015].
<!-- As a result, we expect a moderate relationship. -->

H1: People who are more concerned about their privacy are less likely to communicate actively on a website.

@@

Although privacy calculus implies that people explicitly compare benefits and drawbacks before communicating online, research has neglected this aspect [@knijnenburgDeathPrivacyCalculus2017].
Only observing that privacy concerns or expected gratifications and communication online are related does not prove that an explicit weighing process took place [@knijnenburgDeathPrivacyCalculus2017]. 
Instead, we need to analyze if and by how much people actively deliberate about their privacy, and whether doing so influences their decision to communicate.
<!-- We can understand the privacy calculus from two perspectives. -->
<!-- First, is the communication behavior aligned with people's privacy concerns and expected benefits?  -->
<!-- Second, is the communication process implicit or explicit? -->
<!-- Here, I suggest that the privacy calculus should be discussed in light of dual process theories, which state that people either deliberately, explicitly, and centrally take decisions, or instead do so automatically, implicitly, and peripherally [@kahnemanThinkingFastSlow2011; @pettyCommunicationPersuasionCentral1986].  -->
<!-- Accordingly, privacy calculus would assume that people, when it comes to disclosing, engage in a central processing.  -->
Self-disclosure theory [@omarzuDisclosureDecisionModel2000;  @altmanPrivacyConceptualAnalysis1976] suggests that if the benefits to be attained from communication are attractive, deciding whether or not to communicate is a "conscious and deliberative process"  [@omarzuDisclosureDecisionModel2000, p. 183].
I hence introduce and investigate a novel concept termed _privacy deliberation_.
Privacy deliberation captures the extent to which individuals explicitly compare risks and benefits before communicating with others.

How could deliberating about one's privacy affect communication?
On the one hand, it could reduce subsequent communication.
Refraining from communication---the primary means of connecting with others [@altmanPrivacyConceptualAnalysis1976]---likely requires some active and deliberate restraint [@omarzuDisclosureDecisionModel2000]. 
This is especially true for social media, which are designed to foster communication and participation [@ellisonSocialNetworkSite2015; @masurBehavioralContagionSocial2021].
Actively thinking about whether communicating is worthwhile might be the first step not to participate.
On the other hand, deliberating about privacy might also increase communication.
In light of the benefits mentioned above, would it not make sense to conclude that participation is beneficial, thereby fostering communication [@krasnovaOnlineSocialNetworks2010]?
<!-- I therefore formulate the following research question: -->
<!-- Deliberation could represent some kind of inner consent, providing additional affirmation. -->
<!-- Alternatively, it could be that deliberation functions as a moderator. -->
<!-- For example, if people actively deliberate about whether or not to disclose, this might reinforce the effect of concerns or gratifications. -->
<!-- Reflecting about the pros and cons of communication might concerns and gratifications more salient.  -->
<!-- Alternatively, it could also be that deliberating decreases the effects, for example because apparent gratifications are considered more critically, and maybe loose their appeal. -->

RQ1: Do people who deliberate more actively about their privacy communicate more or less online?

<!-- RQ2: Do people who deliberate more actively whether they should communicate, show larger or smaller relations between concerns, gratifications and communication behavior? -->
This weighing process is not flawless or perfect.
It is therefore useful to understand the privacy calculus from a perspective of _bounded rationality_ [@simonBoundedRationality1990].
Bounded rationality states that "(1) humans are cognitively constrained; (2) these constraints impact decision making; and (3) difficult problems reveal the constraints and highlight their significance." [@bendorBoundedRationality2015, p. 1303].
Although human behavior is partly irrational, bounded rationality does not state that it is completely irrational [@gigerenzerBoundedRationalityAdaptive2002].
Rationality is a continuum. 
<!-- Instead, it makes more sense to consider rationality a continuum. -->
<!-- Humans are still trying to optimize the outcomes of their behavior according to their own best interests or values. -->
<!-- It is only that their capacity to do so is bounded. -->
In the context of online privacy, for example, rationality is impeded by information asymmetries, presence bias, intangibility, illusory control, or herding [@acquistiSecretsLikesDrive2020].
It follows that to provide a more complete picture, additional factors next to gratifications and concerns should also explain communication.
Factors less focused on rationality, but more on socio-psychological aspects.
<!-- Communication online often takes place in situations where information is limited or obscure. -->

@@

In situations where people lack experience or competence, a relevant variable to understand behavior is _trust_.
Especially in online contexts, users often cannot control the environment or the way their information is handled [@acquistiSecretsLikesDrive2020; @braunlichLinkingLooseEnds2020].
Trust plays a key role in online communication [@metzgerPrivacyTrustDisclosure2004]. 
Trust either captures "_specific_ beliefs dealing primarily with the integrity, benevolence, and ability of another party" [@gefenTrustTAMOnline2003, p. 55, emphasis added] or a "_general_ belief that another party can be trusted" [@gefenTrustTAMOnline2003, p.55, emphasis added]. 
<!-- Whereas specific trust focuses on the causes of trust, general trust emphasizes the experience of trust. -->
In online contexts, there are different targets of trust, including (a) the information system, (b) the provider, (c) the Internet, and (d) the community of other users [@sollnerWhyDifferentTrust2016]. 
People who put more trust in the providers of networks, for example, disclose more personal information [@liEmpiricalStudiesOnline2011]. 
<!-- @gefenTrustTAMOnline2003 prioritize specific trust (p. 60).  -->
<!-- Because the targets can be largely different, it is often recommended to analyze them individually. -->

@@

How are the privacy calculus and communication affected by the context, the digital infrastructure? 
In analyzing this question, it makes sense not to focus on specific features of particular websites, which change and become obsolete [@foxDistinguishingTechnologiesSocial2017]. 
Instead, it is recommended focusing on the underlying latent structures by analyzing so-called _affordances_ [@ellisonSocialNetworkSite2015; @foxDistinguishingTechnologiesSocial2017].
Affordances are mental representations of how objects might be used.
They emphasize that it is not the _objective features_ that determine behavior, but rather our _subjective perceptions_ [@gibsonEcologicalApproachVisual2015].
<!-- ; as such, they are by definition subjective.  -->
There is an ongoing debate on how to best understand and operationalize affordances [@evansExplicatingAffordancesConceptual2017].
Whereas @evansExplicatingAffordancesConceptual2017 propose three affordances for mediated communication (i.e., anonymity, persistence, and visibility), @foxDistinguishingTechnologiesSocial2017 suggest 10 affordances for SNSs alone (i.e., accessibility, bandwidth, social presence, privacy, network association, personalization, persistence, editability, conversation control, and anonymity).
<!-- The privacy calculus states that both benefits and costs determine behavior. -->
Popularity cues such as like and dislike buttons, which are of interest here, are categorized as "paralinguistic digital affordances" [@carrPredictingThresholdPerceived2018, p. 142].

@@

Paralinguistic digital affordances and specifically popularity cues have been shown to affect behavior [@kramerMasteringChallengeBalancing2020; @masurBehavioralContagionSocial2021; @treptePrivacyCalculusContextualized2020].
Online comments that already have several dislikes are more likely to receive further dislikes [@muchnikSocialInfluenceBias2013]. 
When users disagree with a post, they are more likely to click on a button labeled _respect_ compared to a button labeled _like_ [@stroudRecommendRespectAltering2017].
<!-- In light of their potentially strong negative effects, most major social networking sites did not implement dislike buttons.  -->
<!-- At the time of writing, only a handful of websites have (partially) implemented it (e.g., youtube, stackexchange, or reddit). -->
Popularity cues likely impact the privacy calculus via two underlying theoretical mechanisms [@kramerMasteringChallengeBalancing2020]:
First, the _mere presence_ of popularity cues might affect whether people are willing to disclose. 
Being able to attract likes might motivate users to communicate, while the mere option to receive dislikes might intimidate others.
Second, _actually receiving_ likes or dislikes might additionally affect subsequent behavior, thereby reinforcing the process [@skinnerScienceHumanBehavior2014].

Likes are affirmative and represent the positivity bias of social media [@schreursExposurePositivityBias2022]. 
Receiving a like online is similar to receiving a compliment offline [@carrPredictingThresholdPerceived2018; @sumnerFunctionalApproachFacebook2017].
Like-buttons afford and emphasize a _gain frame_ [@rosoffHeuristicsBiasesCyber2013].
These gains can be garnered only through active participation. 
Because like buttons emphasize positive outcomes, it is likely that concerns decrease. 
In situations where there is more to win, people might also more actively deliberate about whether or not to communicate. 
Dislikes, instead, represent a punishment, introducing a _loss frame_. 
<!-- Although most communication emphasizes positive aspects, the Internet is also replete with spite, envy, and arguments. -->
Websites featuring both like _and_ dislike buttons should therefore be more ambivalent compared to websites without popularity cues.
<!-- Especially in online contexts, short-term gains often outweigh long-term losses [@acquistiSecretsLikesDrive2020]. -->
<!-- Having both like and dislike buttons might still lead to more gratifications and communication. -->
Privacy concerns should not be reduced anymore:
People who are more concerned about their privacy are also more shy and risk averse [@dienlinPsychologyPrivacyAnalyzing2017].
Implementing the dislike button might therefore increase privacy concerns, thereby canceling out the positive effects of the like button. 
<!-- And because there is more at stake, participants should deliberate even more whether or not to disclose.  -->

@@

This manuscript features a [companion website](https://XMtRa.github.io/like_dislike), which includes the data, research materials, analyses scripts, and a reproducible version of this manuscript (see https://XMtRa.github.io/like_dislike).
The hypotheses, sample size, research materials, analyses, and exclusion criteria were preregistered (see https://osf.io/a6tzc/?view_only=5d0ef9fe5e1745878cd1b19273cdf859). 
In some cases, the preregistered approach needed to be changed (see [companion website](https://XMtRa.github.io/like_dislike/preregistration_changes)). 
Analyses that were not preregistered are reported as exploratory analyses.
